---
title: "Multiple Linear Regression Review"
author: "Jack Cunningham"
format: pdf
editor: visual
---

```{r}
data <- read.csv("HoustonChronicle.csv")
head(data)
attach(data)
```

To answer these three questions we use the model:

$$
Y=\beta_0+\beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2
$$

Where Y is the percentage of students repeating first grade, $x_1$ is the percentage of low income students, and $x_2$ is a dummy variable that is equal to 1 when Year is 2004 and 0 when year is 1994.

```{r}
data$year_dummy <- ifelse(Year == 2004, 1,0)
```

```{r}
fit <- lm(X.Repeating.1st.Grade ~ X.Low.income.students + year_dummy + 
            X.Low.income.students*year_dummy, data = data)
summary(fit)
```

a\)

The association between the percentage of low income students and the percentage of students repeating first grade is low and not statistically significant. If we test the hypothesis:

$$
H_0:\beta_1 =0,H_a:\beta_1 \neq0
$$

The null hypothesis cannot be rejected with a p-value barely over a significance of $\alpha=0.05167$.

The effect size is low as well, $\hat{\beta_1}=0.0608$ means an increase in one percentage of low income students only increases the percentage of students repeating first grade by $0.06\%$.

b\)

To review, when the year is 1994 we have:

$$
Y=\beta_0+\beta_1 x_1
$$

When the year is 2004 we have:

$$
Y=(\beta_0+\beta_2)+(\beta_1+\beta_3)x_1
$$

In order to test if there has been an increase in the percentage of students repeating first grade between 1994-1995 and 2004-2005 let us first test whether there is a difference between the two years. We do this by testing:

$$
H_0:\beta_2=\beta_3=0,H_a:\beta_2 \neq0 \text{ or }
\beta_3 \neq 0$$

We have test statistic:

$$
F=\frac{(RSS_{reduced}-RSS_{full})/(df_{reduced}-df_{full})}{RSS_{full/}/df_{full}}
$$

Where $df_{full} = 61 - 3 -1$ and $df_{reduced}=61 - 1 - 1$.

```{r}
fit_reduced <- lm(X.Repeating.1st.Grade ~ X.Low.income.students, data = data)
summary(fit_reduced)
```

```{r}
df_full = 61 - 3 - 1
df_reduced = 61 - 1 -1
RSS_full = sum(fit$residuals^2)
RSS_reduced = sum(fit_reduced$residuals^2)

F = (RSS_reduced - RSS_full)/(df_reduced - df_full) /
  (RSS_full/df_full)
p_value = 1 - pf(F, df_reduced - df_full, df_full)
cat("F Statistic: ",F, " p value: ", p_value)
```

With a p value of 0.8847324 we can cannot reject the null hypothesis that the reduced model, which disregards year, is as appropriate as the full model.

c\)

To see if there is an interaction effect between year and the percentage of low-income students on the percentage of students repeating first grade we test:

$$
H_0:\beta_3=0,H_1:\beta_3 \neq0
$$

We have test statistic:

$$
T=\frac{\hat{\beta_3}}{SE(\hat{\beta_3})}
$$

```{r}
beta_3_hat <- as.numeric(fit$coefficients[4])
se_beta_3 <- summary(fit)$coefficients[4,2]
df = 61 - 3 - 1
T = beta_3_hat/se_beta_3
p_value <- 2*(1 - pt(T,df))
cat("T statistic: ", T, "p value:", p_value)
```

We cannot reject the null hypothesis that there is no interaction effect between year and the percentage of low-income students on the percentage of students repeating first grade.
